Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-10-31 doc/paper.tex

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\section{Background and Related Work}

@@

To evaluate the effectivness of the integration of LLMs into reinfrocement learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries. Here, we handpick specific environments with varying levels of complexity, to evaluate the effectivness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other. The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game. How difficult an environment is to beat, depends no factors such as shape of the walls, if the agent has to solve subtasks to finish the game, and the type of subtasks encountered. Moreover what makes these environments exceptionally well suited for testing our approach, is that they are sparse, in the context of MiniGrid and BabyAI, this means that the agent will only receive a reward when it finishes the game. It is a well known fact that even the state-of-art RL algorithms struggle in very complex and sparse environments, hence it becomes the perfect arena to see if the high generalization capabilites of Large pre-trained Language Models will lead to better results than using only RL methods.

@@

Our baseline is the \textbf{Proximal Policy Optimization} (PPO), which over the last years has established itself as the state-of-art approach to solving reinforcement learning problems. The main idea behind this algorithms is that the new policy is clipped so that it does not move too quickly and too far from the old policy. I.e. the new policy cannot diverge too moch from the old policy, which leads to exceedingly large updates to policy gradients being avoided, thus also avoiding the problem of vanishing and exploding gradients. Suprisingly, this has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN. For a more detailed discussion of the mathematics and implementation details, see \cite{OpenAIPPO}

@@

In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken. This is done as follows: First, the observation is captioned and given as input to the LLM, which is prompted to suggest an action. Then, all actions that the agent could take given the current state are described in natural language. All possible actions are compared with the suggested action, and given a value which denotes their semantic similarity. These values are joined with the probability values that the normal PPO algorithm outputs and normalized, and a new probability distribution is created, where semantically similar actions are weighted higher than others. We call this the Actor-Advisor-Critic Model.

@@

The agent is trained as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM. This may be done at every timestep, or after some interval of timesteps. In addition to the state observation, the LLM is also asked to respond with a list of recommendations. These recommendations can be proposed actions for the agent to perform, or goal states that the agent should reach. The agent then acts independently of the LLM for one or more timesteps, as dictated by the baseline RL algorithm it is trained with. As it does so, its actions are tracked and encoded to natural text. Alternatively, a new state observation is generated after the timesteps, and also encoded as natural text. Then, the LLM is used as a tool for semantic similarity comparison, comparing either its own formerly proposed actions to the actions performed by the agent, or comparing its proposed goal state to the new state reached by the agent. Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.

@@

For the purpose of our research, we have decided utilize a LLM from the open-source family of language models LLaMA developed by Meta AI, released starting Febraury, 2023. Specifically, we chose the newest of the models named Llama2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama1, and according to \cite{meta} has double the context length. In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call the API-version. The reason for not using any of the larger Llama2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.

@@

We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy integration. All methods use the same hyperparameters for the PPO, which can be found the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM} received for each one. We do this eight times per method, and compare the average results against eachother.

@@

As we can see from figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process. Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards. Furthermore, this is an improvement that comes at very little computational cost, because LLM responses can be cached between runs. Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers. In all methods, we see that the standard deviation is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively). Given that there is no reward given for a partially completed task, this is not unexpected.

@@

@@

@@

In the further development of this and similar algorithms, we also consider it essential to put effort into trying to understand the causality behind certain actions being taken, as is being done in the field of explainable artificial intelligence (XAI). When involving one major ML framework like that of LLMs into RL, we would ideally want to see exactly where and how much the LLM contributes to the RL networks being updated. This would, in turn, help in the work to mitigate whatever biases and variance that may be added to the RL framework from integrating the LLM.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\section{Background and Related Work}
\label{background}

@@

To evaluate the effectiveness of the integration of LLMs into Reinforcement Learning algorithms, we use the \textbf{MiniGrid} and \textbf{BabyAI} libraries. Here, we handpick specific environments with varying levels of complexity, to evaluate the effectiveness of our proposed techniques against a standard Actor-Critic \ref{Proximal policy optimzation}, and against each other. The environments selected are all based in a small sized grid-world, where the agent is surrounded by walls, and has to perform at least one task to complete the game. How difficult an environment is to beat depends on factors such as the shape of the walls, whether the agent has to solve subtasks to finish the game, and the types of subtasks encountered. Moreover, what makes these environments especially well suited for testing our approach is that they are sparse in the context of MiniGrid and BabyAI. This means that the agent will only receive a reward when it finishes the game. It is a well known fact that even state-of-art RL algorithms struggle in very complex and sparse environments. Therefore, it becomes the perfect arena to see if the high generalization capabilites of pre-trained Large Language Models will lead to better results than using only RL methods.

@@

Our baseline is the \textbf{Proximal Policy Optimization} (PPO) algorithm, which has been established as a state-of-art approach to solving reinforcement learning problems over the last few years. The main idea behind this algorithm is that the new policy is clipped so that it does not move too quickly and too far from the old policy, i.e. the new policy cannot diverge too much from the old policy. This helps to avoid exceedingly large updates to policy gradients, thus also avoiding the problem of vanishing and exploding gradients. This has elevated the ability of RL-techniques to tackle sparse environments, and outperform older approches such as DQN. For a more detailed discussion of the mathematics and implementation details, see \cite{OpenAIPPO}

@@

In this case, the LLM is integrated as part of the PPO algorithm, directly influencing the actions taken. This is done as follows: First, the observation is captioned and given as input to the LLM, which is prompted to suggest an action. Then, all actions that the agent could take given the current state are described in natural language. All possible actions are compared with the suggested action, and given a value which denotes their semantic similarity. These values are joined with the logits of the actions, thus influencing the probability weights of each action being taken so that semantically similar actions are weighted higher than others. We call this the Actor-Advisor-Critic Model.

@@

The agent is trained as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM. This may be done at every timestep, or after some interval of timesteps. In addition to the state observation, the LLM is also asked to respond with a list of recommended actions for the agent to perform. The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with. As it does so, its chosen action is encoded to natural text. Then, the LLM is used as a tool for semantic similarity comparison, comparing its own formerly proposed action to the action performed by the agent. Finally, high semantic similarity gives an additional reward which is implemented in the reward function of the agent.

@@

For the purpose of our research, we have decided utilize a LLM from the Llama open-source family of language models developed by Meta AI, first released in Febraury, 2023. Specifically, we chose the newest of the models named Llama 2 \cite{llama}, which has been trained using 40\% more data than its predecessor Llama 1, and according to \cite{meta} has double the context length. In the experiments presented below, we use the smallest version of Llama2, with 7 billion parameters, which we download to run locally and reduce the bottleneck effect of having to call the API-version. The reason for not using any of the larger Llama 2 models is simply the hardware limitation we face, with the 7B model requiring 19GB of VRAM to be run.

@@

We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy optimization ***CONSIDER ALSO TESTING REWARD SHAPER + CONSIGLIERE TOGETHER***. All methods use the same hyperparameters for the PPO, which can be found in the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM in reward shaping} received for each one. We do this eight times per method, and compare the average results against each other.

@@

As we can see from Figure \ref{doorkeyresults}, LLM integration gives a noticable improvement to the performance of the model, not just in the early stages of training, but throughout the entire training process. Averaging over all episodes of all eight runs per model, we find that LLM reward shaping receives $24\%$ more reward per episode than PPO trained without LLM rewards. Furthermore, this is an improvement that comes at very little computational cost, because LLM responses can be cached between runs. Of the eight runs on this environment using LLM reward shaping, the last three runs never queried the LLM at all, only relying on the cache of previous answers. In all methods, we see that the standard deviation is very large ($0.28$ and $0.31$ for LLM reward shaping and baseline, respectively). Given that there is no reward given for a partially completed task, this is not unexpected.

We also notice a significant performance increase when using LLM policy optimization, even somewhat more than from reward shaping. Averaging over all episodes of all runs, we observe a NUMBER\% increase in reward per episode compared to the PPO baseline, and a NUMBER\% increase when compared to LLM policy shaping. The standard deviation when using LLM policy optimization is also similarly large to PPO and reward shaping (NUMBER). By caching LLM responses similarly to when performing reward shaping, this method too adds very little computational cost to the training in total.

***ADD PARAGRAPH ABOUT USING BOTH TOGETHER*** However, the results from both of these methods are eclipsed by the results from combining LLM reward shaping and policy optimization together. Averaging over 8 runs, we observe a NUMBER\% increase in reward from the environment compared to PPO alone, NUMBER\% higher than reward shaping and NUMBER\%higher than policy optimization, with a similar standard deviation of NUMBER.

@@

\subsection{Unlock Pickup environment}

The Unlock Pickup environment is an even more complex environment than DoorKey, thus giving rewards to the agent even more sparsely. As may be anticipated, this added complexity to the state space makes the LLM contribution even more significant to the RL performance.

Compared to the baseline PPO when averaged per episode over all 8 runs, LLM reward shaping gave a NUMBER\% increase in reward from the environment, with a standard deviation of NUMBER (PPO alone had a standard deviation of NUMBER). LLM policy optimization gave a NUMBER\% increase compared to the PPO baseline and a NUMBER\% increase compared to reward shaping. Also here giving the highest average reward per episode, the combination of reward shaping and policy optimization gave NUMBER\% more reward compared to the PPO baseline, NUMBER\% more than reward shaping and NUMBER\% higher than policy optimization.

@@

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. In the Minigrid environments we have tested on, DoorKey and Pickup Unlock, we have demonstrated so. This has been done both through the LLM shaping the RL reward function, and through the LLM influencing the action probabilities of the RL agent, as well as through combining these two methodologies.

For the DoorKey environment, we found that all three methods surpassed the performance of our PPO baseline.

@@

In the further development of this and similar algorithms, we also consider it essential to put effort into trying to understand the causality behind certain actions being taken, as is being done in the field of explainable artificial intelligence (XAI). When involving a major ML framework like that of LLMs into RL, we would ideally want to see exactly where and how much the LLM contributes to the RL networks being updated. This would, in turn, help in the work to mitigate whatever biases and variance that may be added to the RL framework from integrating the LLM.
